biological data preperation

### 2018-09-05 src/biological_data_prep.Rmd
trimmed by date, finished institutions, removed Icelandic data

#Overview
Preperation of the capelin (*Mallotus villosus*) data obtained from [OBIS](www.iobis.org). This database shows the global occurance of capelin. 
Built with 'r getRversion()'.

#Package dependencies
You can load them using the following code which uses a function called [ipak](https://gist.github.com/stevenworthington/3178163). 
Note this function checks to see if the packages are installed first.
Th
```{r pre-install packages, message=FALSE}
packages <- c("packrat", "rworldmap")
source("../src/ipak.R")
ipak(packages)
suppressMessages(ipak)
```

#Raw Data Clean
Load the raw data, then check use head() to check it looks ok, then names() to see all the column names
```{r load raw data}
rawdata <- read.csv("../data/bio/capelin_obis_raw.csv")
head(rawdata) 
names(rawdata)
```

@@

The dataset is global and needs to be trimmed to contain just the area of interest. 
Plot all the occurance points
```{r map all occurrence points}
library(rworldmap)
map <- getMap(resolution = "low") #creates an object called map at medium resoultion
plot(map, xlim = c(-180, -44), ylim =c(0,80), asp = 1) #the x and y lim are the long-lat bounds of the map
points(tidydata$decimalLongitude, tidydata$decimalLatitude) #this adds points to the mapet", xlab = "Longitude", ylab = "Latitude")
```
now extract just the Atlantic Canada data.
The lon-lat study area (which captures the whole Atlantic Canadian EEZ up to the north of Resolution Island in the Hudson Strait/ Davis Strait) is -71.0,-43.0,62.0,38.0 
```{r map only atlantic canadian occurrences}
tidydata <- tidydata[tidydata$decimalLongitude > -71, ]
tidydata <- tidydata[tidydata$decimalLongitude < -43, ]
tidydata <- tidydata[tidydata$decimalLatitude > 38, ]
plot(map, xlim = c(-180, -44), ylim =c(0,80), asp = 1) #re-map - this is just atl. canada (after checked subsetting is good)
points(tidydata$decimalLongitude, tidydata$decimalLatitude)
```
ok looks good...save what you have so far to a .csv
```{r}
write.csv(tidydata, file = "../data//bio/capelin_obis_tidy.csv", row.names = FALSE)
```

>>>>

# Overview
Preperation of the capelin (*Mallotus villosus*) data obtained from [OBIS](www.iobis.org). This database shows the global occurance of capelin. 
Built with 'r getRversion()'.

# Package dependencies
You can load them using the following code which uses a function called [ipak](https://gist.github.com/stevenworthington/3178163). 
Note this function checks to see if the packages are installed first.
Th
```{r pre-install packages, message=FALSE}
packages <- c("packrat", "rworldmap", "dplyr" "plyr")
source("../src/ipak.R")
ipak(packages)
suppressMessages(ipak)
```

# Raw Data Clean
Load the raw data, then check use head() to check it looks ok, then names() to see all the column names
```{r load raw data}
rawdata <- read.csv("../data/bio/capelin_obis_raw.csv")
head(rawdata) 
names(rawdata)
```

@@

The dataset is global and needs to be trimmed to contain just the area of interest. 
Plot all the occurance points
```{r map all occurrence points}
library(rworldmap)
map <- getMap(resolution = "low") #creates an object called map at low resoultion
plot(map, xlim = c(-180, -44), ylim =c(0,80), asp = 1) #the x and y lim are the long-lat bounds of the map
points(tidydata$decimalLongitude, tidydata$decimalLatitude) #this adds points to the mapet", xlab = "Longitude", ylab = "Latitude")
```
now extract just the Atlantic Canada data.
The lon-lat study area (which captures the whole Atlantic Canadian EEZ up to the north of Resolution Island in the Hudson Strait/ Davis Strait) is -71.0,-43.0,62.0,38.0 
```{r map only atlantic canadian occurrences}
tidydata <- tidydata[tidydata$decimalLongitude > -71, ]
tidydata <- tidydata[tidydata$decimalLongitude < -43, ]
tidydata <- tidydata[tidydata$decimalLatitude > 38, ]
plot(map, xlim = c(-180, -44), ylim =c(0,80), asp = 1) 
points(tidydata$decimalLongitude, tidydata$decimalLatitude)
```
ok looks good...save what you have so far to a .csv
```{r save tidydata to csv}
write.csv(tidydata, file = "../data/bio/capelin_obis_tidy.csv", row.names = FALSE)
```

# Column removal
So there are many columns that you don't need.. you want to keep
id
decimalLatitude
decimalLongitude
datecollected
institutioncode
individualcount
resname
originalscientificname

```{r select colums needed}
tidydata <- subset(tidydata, select = c(id, decimalLatitude, decimalLongitude, datecollected, institutioncode, individualcount, resname, originalscientificname))
head(tidydata)
```

# remove data without month/year info

count NAs in the dataset. Mainly you are interested in the datecollected column (ignore individual count - you know there is a lot missing)
```{r}
missingdata <- apply(tidydata, 2, function (x) sum(is.na(x)))
missingdata
```
hmm interesting - you have 771 columns from year month and day with missing data, but datacollected has 0. There may be a formatting issue.
To take a closer look, subset all rows where year == na

```{r}
missingdatarows <- tidydata[is.na(tidydata$year), ]
head(missingdatarows)
```
so the NAs in the year column relate to no data (without a na value attached). Remove all rows where year == na
```{r}
tidydata <- tidydata[!is.na(tidydata$year), ]
missingdata2 <- apply(tidydata, 2, function (x) sum(is.na(x)))
missingdata2
```

# Month year column creation
make columns for the years and month based on the datecollected column
```{r new yymmdd columns}
tidydata$year <- format(as.Date(tidydata$datecollected), "%Y") #created a new column called year
tidydata$month <- format(as.Date(tidydata$datecollected), "%m") #created a new column called month
tidydata$day <- format(as.Date(tidydata$datecollected), "%d") #created a new column called day 
head(tidydata)
```

the year, month, day columns are characters Change to numeric
```{r new yymmdd columns 2}
tidydata$year <- as.numeric(tidydata$year)
tidydata$month <- as.numeric(tidydata$month)
tidydata$day <- as.numeric(tidydata$day)
head(tidydata)
```

# Remove any duplicate entries
The dplyr package's distinct function removes any duplicate rows (based on all columns EXCEPT id as this is always going to be unique, and year month day as this is generated from datecollected)

```{r}
library("dplyr")
tidydata <- distinct(tidydata, decimalLatitude, decimalLongitude, datecollected, institutioncode, institutioncode, individualcount, resname, originalscientificname)
```

# Institutions
Lets take a look at the institutioncode to make sure the names make sense...
```{r}
tidydata$institutioncode <- as.character(tidydata$institutioncode)
whosampled <- unique(tidydata$institutioncode) #makes a variable of institutions
whosampledordered <- sort(whosampled) #just puts the list in 
whosampledordered
```
Ok so a number of these names actually represent the same insitiution (e.g. [DFO Regions](http://www.dfo-mpo.gc.ca/regions/index-eng.htm). Some details [here](ftp://ftp.aoml.noaa.gov/od/pub/library/methods-branton.pdf) and the first is no institution (you might be able to figure it out though). 
The merges needed are as follows
"Bedford Institute of Oceanography (BIO)" & "BIO" & "DFO-SF" == DFOMTMS 
"DFO-IML" & "IML-MLI" & "IML_MLI" & "DFO-NG" == DFOQC 
"DFO-NFLD" & "North Atlantic Fisheries Centre" == "DFONL" 
"NMFS-NEFSC" & "NOAA, NMFS, Northeast Fisheries Science Center" == NEFSC
"The Atlantic reference Centre" & "The Atlantic Reference Centre (ARC)" == ARC 
"DFO-SG" & "DFO Gulf Fisheries Centre" == DFOGULF

and some you can just make shorter
"Northwest Atlantic Fisheries Organization (NAFO)" == NAFO
"DFO-ISDM" == DFOISDM
"Maine DMR" == MaineDMR
```{r consistent institutioncodes}
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "Bedford Institute of Oceanography (BIO)", "DFOMTMS")
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "BIO", "DFOMTMS")
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "DFO-SF", "DFOMTMS")
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "DFO-IML", "DFOQC")
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "IML-MLI", "DFOQC")
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "DFO-NG", "DFOQC")  
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "IML_MLI", "DFOQC")
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "DFO-NFLD", "DFONL")
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "North Atlantic Fisheries Centre", "DFONL")
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "NMFS-NEFSC", "NEFSC")
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "NOAA, NMFS, Northeast Fisheries Science Center", "NEFSC")
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "The Atlantic Reference Centre (ARC)", "ARC")
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "The Atlantic reference Centre", "ARC")
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "DFO-SG", "DFOGulf")
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "DFO Gulf Fisheries Centre", "DFOGulf")
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "Northwest Atlantic Fisheries Organization (NAFO)", "NAFO")
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "DFO-Central and Arctic, Freshwater Institute (FWI)", "DFOCENARC")
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "DFO-ISDM", "DFOISDM")
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "Maine DMR", "MaineDMR")

#recheck
tidydata$institutioncode <- as.character(tidydata$institutioncode)
whosampled <- unique(tidydata$institutioncode) #makes a variable of institutions
whosampledordered <- sort(whosampled) #just puts the list in 
whosampledordered
```

Now look at the missing institution. Use resname to see if it helps...
```{r check missing institutioncode}
missinginst <- subset(tidydata, institutioncode == "")
missinginst
```
ok now check for unique resname...
```{r ID missing institutioncode}
missinginst$resname <- as.character(missinginst$resname)
missinginstunique <- unique(missinginst$resname) #makes a variable of institutions
missinginstuniquedordered <- sort(missinginstunique) #just puts the list in 
missinginstuniquedordered <- as.character(missinginstuniquedordered)
missinginstuniquedordered
```

Ok soo looking at the [OBIS metadata](http://www.iobis.org/explore/#/dataset/4356), the institution is the Conservation of Arctic Flora and Fauna (CAFF)
```{r fix missing institutioncode}
tidydata$institutioncode <- replace(tidydata$institutioncode, tidydata$institutioncode == "", "CAFF")
whosampled <- unique(tidydata$institutioncode) #makes a variable of institutions
whosampledordered <- sort(whosampled) #just puts the list in 
whosampledordered
```

Now lets check the number of records and spatial-temporal distribution of the observations by institution code to make sure none are dodgy

first a table of how many observations each instituioncode has
```{r institution code analysis - count}
library(plyr) #to use the count function
obs_by_ins <- count(tidydata, "institutioncode")
obs_by_ins
write.csv(obs_by_ins, file = "../output/bio/no_observations_institutioncode.csv")
```
ok so IMR, NAFO, ICES, & USM have very few entries (1, 1, 1, and 3 respectively)

Lets take a look at the spatial breakdown of these institutions.First all points...

```{r}
map2 <- getMap(resolution = "low") #creates an object called map at low resoultion
plot(map2, xlim = c(-71, -43), ylim =c(38, 62), asp = 1, main = "All Occurrences", col = "cornsilk") #the x and y lim are the long-lat bounds of the map
points(tidydata$decimalLongitude, tidydata$decimalLatitude, col = "red") #this adds points to the mapet", xlab = "Longitude", ylab = "Latitude")
dev.copy(png, "../output/bio/all_occurrences.png") #this prints a png of the plot
dev.off() #this turns off the print command
```
Note there is one point up by iceland that you should get rid of (Icelandic population thought to be seperate from Labrador, but it is unclear if this is true or not).

Map the institutioncode == ARC only data...
```{r map obs by institutuion}
ARC_obs <- tidydata[tidydata$institutioncode == "ARC", ]
map2 <- getMap(resolution = "low") #creates an object called map at low resoultion
plot(map2, xlim = c(-71, -43), ylim =c(39, 62), asp = 1, main = "Arc Occurrences", col = "cornsilk") #the x and y lim are the long-lat bounds of the map
points(ARC_obs$decimalLongitude, ARC_obs$decimalLatitude, col = "red") #this adds points to the mapet", xlab = "Longitude", ylab = "Latitude")
dev.copy(png, "../output/bio/ARC_occurrences_all.png") #this prints a png of the plot
dev.off() #this turns off the print command

CAFF_obs <- tidydata[tidydata$institutioncode == "CAFF", ]
map2 <- getMap(resolution = "low") #creates an object called map at low resoultion
plot(map2, xlim = c(-71, -43), ylim =c(39, 62), asp = 1, main = "CAFF Occurrences",  col = "cornsilk") #the x and y lim are the long-lat bounds of the map
points(CAFF_obs$decimalLongitude, CAFF_obs$decimalLatitude, col = "red") #this adds points to the mapet", xlab = "Longitude", ylab = "Latitude")
dev.copy(png, "../output/bio/CAFF_occurrences_all.png") #this prints a png of the plot
dev.off() #this turns off the print command

DFOCENARC_obs <- tidydata[tidydata$institutioncode == "DFOCENARC", ]
map2 <- getMap(resolution = "low") #creates an object called map at low resoultion
plot(map2, xlim = c(-71, -43), ylim =c(39, 62), asp = 1, main = "DFO Central & Arctic Occurrences",  col = "cornsilk") #the x and y lim are the long-lat bounds of the map
points(DFOCENARC_obs$decimalLongitude, DFOCENARC_obs$decimalLatitude, col = "red") #this adds points to the mapet", xlab = "Longitude", ylab = "Latitude")
dev.copy(png, "../output/bio/DFOCENARC_occurrences_all.png") #this prints a png of the plot
dev.off() #this turns off the print command

DFOGulf_obs <- tidydata[tidydata$institutioncode == "DFOGulf", ]
map2 <- getMap(resolution = "low") #creates an object called map at low resoultion
plot(map2, xlim = c(-71, -43), ylim =c(39, 62), asp = 1, main = "DFO Gulf Occurrences",  col = "cornsilk") #the x and y lim are the long-lat bounds of the map
points(DFOGulf_obs$decimalLongitude, DFOGulf_obs$decimalLatitude, col = "red") #this adds points to the mapet", xlab = "Longitude", ylab = "Latitude")
dev.copy(png, "../output/bio/DFOGulf_occurrences_all.png") #this prints a png of the plot
dev.off() #this turns off the print command

DFOISDM_obs <- tidydata[tidydata$institutioncode == "DFOISDM", ]
map2 <- getMap(resolution = "low") #creates an object called map at low resoultion
plot(map2, xlim = c(-71, -43), ylim =c(39, 62), asp = 1, main = "DFO ISDM Occurrences",  col = "cornsilk") #the x and y lim are the long-lat bounds of the map
points(DFOISDM_obs$decimalLongitude, DFOISDM_obs$decimalLatitude, col = "red") #this adds points to the mapet", xlab = "Longitude", ylab = "Latitude")
dev.copy(png, "../output/bio/DFOISDM_occurrences_all.png") #this prints a png of the plot
dev.off() #this turns off the print command

DFOMTMS_obs <- tidydata[tidydata$institutioncode == "DFOMTMS", ]
map2 <- getMap(resolution = "low") #creates an object called map at low resoultion
plot(map2, xlim = c(-71, -43), ylim =c(39, 62), asp = 1, main = "DFO Maritimes Occurrences",  col = "cornsilk") #the x and y lim are the long-lat bounds of the map
points(DFOMTMS_obs$decimalLongitude, DFOMTMS_obs$decimalLatitude, col = "red") #this adds points to the mapet", xlab = "Longitude", ylab = "Latitude")
dev.copy(png, "../output/bio/DFOMTMS_occurrences_all.png") #this prints a png of the plot
dev.off() #this turns off the print command

DFONL_obs <- tidydata[tidydata$institutioncode == "DFONL", ]
map2 <- getMap(resolution = "low") #creates an object called map at low resoultion
plot(map2, xlim = c(-71, -43), ylim =c(39, 62), asp = 1, main = "DFO Newfouundland & Labrador Occurrences",  col = "cornsilk") #the x and y lim are the long-lat bounds of the map
points(DFONL_obs$decimalLongitude, DFONL_obs$decimalLatitude, col = "red") #this adds points to the mapet", xlab = "Longitude", ylab = "Latitude")
dev.copy(png, "../output/bio/DFONL_occurrences_all.png") #this prints a png of the plot
dev.off() #this turns off the print command

DFOQC_obs <- tidydata[tidydata$institutioncode == "DFOQC", ]
map2 <- getMap(resolution = "low") #creates an object called map at low resoultion
plot(map2, xlim = c(-71, -43), ylim =c(39, 62), asp = 1, main = "DFO Quebec Occurrences",  col = "cornsilk") #the x and y lim are the long-lat bounds of the map
points(DFOQC_obs$decimalLongitude, DFOQC_obs$decimalLatitude, col = "red") #this adds points to the mapet", xlab = "Longitude", ylab = "Latitude")
dev.copy(png, "../output/bio/DFOQC_occurrences_all.png") #this prints a png of the plot
dev.off() #this turns off the print command

ICES_obs <- tidydata[tidydata$institutioncode == "ICES", ]
map2 <- getMap(resolution = "low") #creates an object called map at low resoultion
plot(map2, xlim = c(-71, -43), ylim =c(39, 62), asp = 1, main = "ICES Occurrences",  col = "cornsilk") #the x and y lim are the long-lat bounds of the map
points(ICES_obs$decimalLongitude, ICES_obs$decimalLatitude, col = "red") #this adds points to the mapet", xlab = "Longitude", ylab = "Latitude")
dev.copy(png, "../output/bio/ICES_occurrences_all.png") #this prints a png of the plot
dev.off() #this turns off the print command

IMR_obs <- tidydata[tidydata$institutioncode == "IMR", ]
map2 <- getMap(resolution = "low") #creates an object called map at low resoultion
plot(map2, xlim = c(-71, -43), ylim =c(39, 62), asp = 1, main = "IMR Occurrences",  col = "cornsilk") #the x and y lim are the long-lat bounds of the map
points(IMR_obs$decimalLongitude, IMR_obs$decimalLatitude, col = "red") #this adds points to the mapet", xlab = "Longitude", ylab = "Latitude")
dev.copy(png, "../output/bio/IMR_occurrences_all.png") #this prints a png of the plot
dev.off() #this turns off the print command

MaineDMR_obs <- tidydata[tidydata$institutioncode == "MaineDMR", ]
map2 <- getMap(resolution = "low") #creates an object called map at low resoultion
plot(map2, xlim = c(-71, -43), ylim =c(39, 62), asp = 1, main = "Maine DMR Occurrences",  col = "cornsilk") #the x and y lim are the long-lat bounds of the map
points(MaineDMR_obs$decimalLongitude, MaineDMR_obs$decimalLatitude, col = "red") #this adds points to the mapet", xlab = "Longitude", ylab = "Latitude")
dev.copy(png, "../output/bio/MaineDMR_occurrences_all.png") #this prints a png of the plot
dev.off() #this turns off the print command

NAFO_obs <- tidydata[tidydata$institutioncode == "NAFO", ]
map2 <- getMap(resolution = "low") #creates an object called map at low resoultion
plot(map2, xlim = c(-71, -43), ylim =c(39, 62), asp = 1, main = "NAFO Occurrences",  col = "cornsilk") #the x and y lim are the long-lat bounds of the map
points(NAFO_obs$decimalLongitude, NAFO_obs$decimalLatitude, col = "red") #this adds points to the mapet", xlab = "Longitude", ylab = "Latitude")
dev.copy(png, "../output/bio/NAFO_occurrences_all.png") #this prints a png of the plot
dev.off() #this turns off the print command

NEFSC_obs <- tidydata[tidydata$institutioncode == "NEFSC", ]
map2 <- getMap(resolution = "low") #creates an object called map at low resoultion
plot(map2, xlim = c(-71, -43), ylim =c(39, 62), asp = 1, main = "NEFSC Occurrences",  col = "cornsilk") #the x and y lim are the long-lat bounds of the map
points(NEFSC_obs$decimalLongitude, NEFSC_obs$decimalLatitude, col = "red") #this adds points to the mapet", xlab = "Longitude", ylab = "Latitude")
dev.copy(png, "../output/bio/NEFSC_occurrences_all.png") #this prints a png of the plot
dev.off() #this turns off the print command

ROM_obs <- tidydata[tidydata$institutioncode == "ROM", ]
map2 <- getMap(resolution = "low") #creates an object called map at low resoultion
plot(map2, xlim = c(-71, -43), ylim =c(39, 62), asp = 1, main = "Royal Ontario Museum Occurrences",  col = "cornsilk") #the x and y lim are the long-lat bounds of the map
points(ROM_obs$decimalLongitude, ROM_obs$decimalLatitude, col = "red") #this adds points to the mapet", xlab = "Longitude", ylab = "Latitude")
dev.copy(png, "../output/bio/ROM_occurrences_all.png") #this prints a png of the plot
dev.off() #this turns off the print command

USNM_obs <- tidydata[tidydata$institutioncode == "USNM", ]
map2 <- getMap(resolution = "low") #creates an object called map at low resoultion
plot(map2, xlim = c(-71, -43), ylim =c(39, 62), asp = 1, main = "National Museum of Natural History Occurrences",  col = "cornsilk") #the x and y lim are the long-lat bounds of the map
points(USNM_obs$decimalLongitude, USNM_obs$decimalLatitude, col = "red") #this adds points to the mapet", xlab = "Longitude", ylab = "Latitude")
dev.copy(png, "../output/bio/USNM_occurrences_all.png") #this prints a png of the plot
dev.off() #this turns off the print command
```

Looking at the maps, the ROM, NAFO, IMR, ICES have data next to iceland.
Since NAFO, IMR, and ICES only have one data point each, you can remove these observations easily from the record...
```{r}
tidydata <- tidydata[!tidydata$institutioncode == "NAFO", ]
tidydata <- tidydata[!tidydata$institutioncode == "IMR", ]
tidydata <- tidydata[!tidydata$institutioncode == "ICES", ]
```
Now for  ROM there are more data points...

write what you have so far to a csv
```{r}
write.csv(tidydata, file = "../data/bio/capelin_obis_tidy.csv", row.names = FALSE)
```
